---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r srr-tags, eval = FALSE, echo = FALSE}
#' @srrstats {G1.2} Contains project status badge.
#' @srrstats {G1.4,G1.4a} Package uses roxygen2 for documentation.
#' @srrstats {G2.0, G2.0a, G2.1, G2.1a, G2.2, G2.4, G2.4a, G2.4b, G2.4c, G2.6} 
#' Input types and shapes are tested and checked with autotest and converted 
#' explicitly when necessary.
#' 
#' @srrstats {G2.3, G2.3a, G2.3b} match.arg and tolower are used where 
#' applicable.
#' @srrstats {G1.0, G1.1, G1.3, G1.4, G1.4a, G1.5, G1.6} General 
#' documentation, addressed by the paper vignette and the corresponding R 
#' Journal paper.
#' @srrstats {G2.4, G2.4a, G2.4b, G2.4c, G2.6} Explicit conversions are used 
#' where necessary.
#' 
#' @srrstats {G2.14, G2.14a, G2.14b, G2.14c, G2.15, G2.16} Missing observations 
#' (y) are handled automatically as per SSM theory, whereas missing values are 
#' not allowed elsewhere. Inputing or ignoring them does not make sense in time 
#' series context.
#'
#' @srrstats {G3.0} No floating point equality comparisons are made.
#'
#' @srrstats {G5.4, G5.4a, G5.4b, G5.4c, G5.5, G5.6, G5.6a, G5.6b, G5.7} The 
#' algorithms work correctly as per Vihola, Helske, Franks (2020) 
#' (all simulations were implemented with the bssm package) and Helske 
#' and Vihola (2021). Full replication of the results would take days/weeks
#' (but see also bsm_ng, negbin_series and several testthat tests).
#'
#' @srrstats {G5.8, G5.8a, G5.8b, G5.8c, G5.8d} Tested with autotest and the 
#' testthat tests.
#' @srrstats {G5.9, G5.9a, G5.9b} Tested with autotest and the testthat tests.
#'
#' @srrstats {BS1.0, BS1.1, BS1.2, BS1.2a, BS1.2b, BS1.3b} Addressed in the 
#' models.R, run_mcmc.R, in vignettes and in the R Journal paper.
#'
#' @srrstats {BS2.1, BS2.1a, BS2.6} Tested and demonstrated by autotest and 
#' package examples/tests.
#' @srrstats {BS7.4, BS7.4a} The scales do not matter (in terms of runtime) 
#' in random walk Metropolis nor in particle filters, as long as numerical 
#' issues are not encountered

```

# bssm

<!-- badges: start -->
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/helske/bssm/workflows/R-CMD-check/badge.svg)](https://github.com/helske/bssm/actions)
[![Codecov test coverage](https://codecov.io/gh/helske/bssm/branch/master/graph/badge.svg)](https://codecov.io/gh/helske/bssm?branch=master)
[![cran version](http://www.r-pkg.org/badges/version/bssm)](http://cran.r-project.org/package=bssm)
[![downloads](http://cranlogs.r-pkg.org/badges/bssm)](http://cranlogs.r-pkg.org/badges/bssm)

<!-- badges: end -->

The `bssm` R package provides efficient methods for Bayesian inference of state 
space models via particle Markov chain Monte Carlo and importance sampling type 
weighted MCMC. 
Currently Gaussian, Poisson, binomial, negative binomial, and Gamma observation 
densities with linear-Gaussian state dynamics, as well as general non-linear 
Gaussian models and discretely observed latent diffusion processes are 
supported.

For details, see 

* [The bssm paper on ArXiv](https://arxiv.org/abs/2101.08492) (to appear in R Journal), 
* [Package vignettes at CRAN](https://cran.r-project.org/web/packages/bssm/index.html) 
* Paper on [Importance sampling type estimators based on approximate marginal Markov chain Monte Carlo](https://onlinelibrary.wiley.com/doi/abs/10.1111/sjos.12492)

There are also couple posters and a talk related to IS-correction methodology and bssm package: 

* [UseR!2021 talk slides](https://jounihelske.netlify.app/talk/user2021/)    
* [SMC 2017 workshop: Accelerating MCMC with an approximation ](http://users.jyu.fi/~jovetale/posters/SMC2017)
* [UseR!2017: Bayesian non-Gaussian state space models in R](http://users.jyu.fi/~jovetale/posters/user2017.pdf)

The `bssm` package was originally developed with the support of Academy of 
Finland grants 284513, 312605, and 311877. Current development is focused on 
increased usability. For recent changes, see NEWS file.

## Installation

You can install the released version of bssm from [CRAN](https://CRAN.R-project.org) with:

```{r, eval=FALSE}
install.packages("bssm")
```

And the development version from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("helske/bssm")
```
Or from R-universe with 

```{r, eval = FALSE}
install.packages("bssm", repos = "https://helske.r-universe.dev")
```

## Example

Consider the daily air quality measurements in New Your from May to September 1973, available in the `datasets` package. Let's try to predict the missing ozone levels by simple linear-Gaussian local linear trend model with temperature and wind as explanatory variables (missing response variables are handled naturally in the state space modelling framework, however no missing values in covariates are normally allowed);

```{r example}
library("bssm")
library("dplyr")
library("ggplot2")
set.seed(1)

data("airquality", package = "datasets")

# Covariates as matrix. For complex cases, check out as_bssm function
xreg <- airquality %>% select(Wind, Temp) %>% as.matrix()

model <- bsm_lg(airquality$Ozone,
  xreg = xreg,  
  # Define priors, see ?bssm_prior
  # Initial value followed by parameters of the prior distribution
  beta = normal_prior(rep(0, ncol(xreg)), 0, 1),
  sd_y = gamma_prior(1, 2, 0.01),
  sd_level = gamma_prior(1, 2, 0.01), 
  sd_slope = gamma_prior(1, 2, 0.01))

fit <- run_mcmc(model, iter = 20000, burnin = 5000)
fit

obs <- data.frame(Time = 1:nrow(airquality),
  Ozone = airquality$Ozone) %>% filter(!is.na(Ozone))

pred <- fitted(fit, model)
pred %>%
  ggplot(aes(x = Time, y = Mean)) + 
  geom_ribbon(aes(ymin = `2.5%`, ymax = `97.5%`), 
    alpha = 0.5, fill = "steelblue") + 
  geom_line() + 
  geom_point(data = obs, 
    aes(x = Time, y = Ozone), colour = "Tomato") +
  theme_bw()

```

Same model but now assuming observations are from Gamma distribution:
```{r gamma-example}

model2 <- bsm_ng(airquality$Ozone,
  xreg = xreg,  
  beta = normal(rep(0, ncol(xreg)), 0, 1),
  distribution = "gamma",
  phi = gamma_prior(1, 2, 0.01),
  sd_level = gamma_prior(1, 2, 0.1), 
  sd_slope = gamma_prior(1, 2, 0.1))

fit2 <- run_mcmc(model2, iter = 20000, burnin = 5000, particles = 10)
fit2
```

Comparison:
```{r compare}
pred2 <- fitted(fit2, model2)

bind_rows(list(Gaussian = pred, Gamma = pred2), .id = "Model") %>%
  ggplot(aes(x = Time, y = Mean)) + 
  geom_ribbon(aes(ymin = `2.5%`, ymax = `97.5%`, fill = Model), 
    alpha = 0.25) + 
  geom_line(aes(colour = Model)) + 
  geom_point(data = obs, 
    aes(x = Time, y = Ozone)) +
  theme_bw()
```


Now let's assume that we also want to use the solar radiation variable as predictor for ozone. As it contains few missing values, we cannot use it directly. As the number of missing time points is very small, simple imputation would likely be acceptable, but let's consider more another approach. For simplicity, the slope terms of the previous models are now omitted, and we focus on the Gaussian case. Let $\mu_t$ be the true solar radiation at time $t$. Now for ozone $O_t$ we assume following model:
$$
\begin{aligned}
O_t &= D_t + \alpha_t + \beta_S \mu_t + \sigma_\epsilon \epsilon_t,\\
\alpha_{t+1} &= \alpha_t + \sigma_\eta\eta_t,\\
\alpha_1 &\sim N(0, 100^2\textrm{I}),
\end{aligned}
$$
where $D_t = \beta X_t$ contains regression terms related to wind and temperature, $\alpha_t$ is the time varying intercept term, and $\beta_S$ is the effect of solar radiation $\mu_t$.

Now for the observed solar radiation $S_t$ we assume 

$$
\begin{aligned}
S_t &= \mu_t\\
\mu_{t+1} &= \mu_t + \sigma_\xi\xi_t,\\
\mu_1 &\sim N(0, 100^2),
\end{aligned}
$$
i.e. we assume as simple random walk for the $\mu$ which we observe without error or not at all (there is no error term in the observation equation $S_t=\mu_t$).

We combine these two models as a bivariate Gaussian model with `ssm_mlg`:

```{r missing-values}
# predictors (not including solar radiation) for ozone
xreg <- airquality %>% select(Wind, Temp) %>% as.matrix()

# Function which outputs new model components given the parameter vector theta
update_fn <- function(theta) {
  D <- rbind(t(xreg %*% theta[1:2]), 1)
  Z <- matrix(c(1, 0, theta[3], 1), 2, 2)
  R <- diag(exp(theta[4:5]))
  H <- diag(c(exp(theta[6]), 0))
  # add third dimension so we have p x n x 1, p x m x 1, p x p x 1 arrays
  dim(Z)[3] <- dim(R)[3] <- dim(H)[3] <- 1
  list(D = D, Z = Z, R = R, H = H)
}

# Function for log-prior density
prior_fn <- function(theta) {
  sum(dnorm(theta[1:3], 0, 10, log = TRUE)) + 
    sum(dgamma(exp(theta[4:6]), 2, 0.01, log = TRUE)) + 
    sum(theta[4:6]) # log-jacobian
}

init_theta <- c(0, 0, 0, log(50), log(5), log(20))
comps <- update_fn(init_theta)

model <- ssm_mlg(y = cbind(Ozone = airquality$Ozone, Solar = airquality$Solar.R),
  Z = comps$Z, D = comps$D, H = comps$H, T = diag(2), R = comps$R, 
  a1 = rep(0, 2), P1 = diag(100, 2), init_theta = init_theta, 
  state_names = c("alpha", "mu"), update_fn = update_fn,
  prior_fn = prior_fn)

fit <- run_mcmc(model, iter = 60000, burnin = 10000)
fit
```

Draw predictions:
```{r bivariate-fig}
pred <- fitted(fit, model)

obs <- data.frame(Time = 1:nrow(airquality),
  Solar = airquality$Solar.R) %>% filter(!is.na(Solar))

pred %>% filter(Variable == "Solar") %>%
  ggplot(aes(x = Time, y = Mean)) + 
  geom_ribbon(aes(ymin = `2.5%`, ymax = `97.5%`), 
    alpha = 0.5, fill = "steelblue") + 
  geom_line() + 
  geom_point(data = obs, 
    aes(x = Time, y = Solar), colour = "Tomato") +
  theme_bw()


obs <- data.frame(Time = 1:nrow(airquality),
  Ozone = airquality$Ozone) %>% filter(!is.na(Ozone))

pred %>% filter(Variable == "Ozone") %>%
  ggplot(aes(x = Time, y = Mean)) + 
  geom_ribbon(aes(ymin = `2.5%`, ymax = `97.5%`), 
    alpha = 0.5, fill = "steelblue") + 
  geom_line() +  
  geom_point(data = obs, 
    aes(x = Time, y = Ozone), colour = "Tomato") +
  theme_bw()
```

