---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r srr-tags, eval = FALSE, echo = FALSE}
#' @srrstatsTODO {G1.2} Contains project status badge.
```
# bssm

<!-- badges: start -->
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![R-CMD-check](https://github.com/helske/bssm/workflows/R-CMD-check/badge.svg)](https://github.com/helske/bssm/actions)
[![Codecov test coverage](https://codecov.io/gh/helske/bssm/branch/master/graph/badge.svg)](https://codecov.io/gh/helske/bssm?branch=master)
[![cran version](http://www.r-pkg.org/badges/version/bssm)](http://cran.r-project.org/package=bssm)
[![downloads](http://cranlogs.r-pkg.org/badges/bssm)](http://cranlogs.r-pkg.org/badges/bssm)
  
<!-- badges: end -->

The `bssm` R package provides efficient methods for Bayesian inference of state 
space models via particle Markov chain Monte Carlo and importance sampling type 
weighted MCMC. 
Currently Gaussian, Poisson, binomial, negative binomial, and Gamma observation 
densities with linear-Gaussian state dynamics, as well as general non-linear 
Gaussian models and discretely observed latent diffusion processes are 
supported.
  
For details, see 
  
  * [the bssm paper on ArXiv](https://arxiv.org/abs/2101.08492), 
  * [package vignettes at CRAN](https://cran.r-project.org/web/packages/bssm/index.html) 
  * Paper on [Importance sampling type estimators based on approximate marginal Markov chain Monte Carlo](https://onlinelibrary.wiley.com/doi/abs/10.1111/sjos.12492)
  
There are also couple posters and a talk related to IS-correction methodology and bssm package: 

  * [UseR!2021 talk slides](https://jounihelske.netlify.app/talk/user2021/)    
  * [SMC 2017 workshop: Accelerating MCMC with an approximation ](http://users.jyu.fi/~jovetale/posters/SMC2017)
  * [UseR!2017: Bayesian non-Gaussian state space models in R](http://users.jyu.fi/~jovetale/posters/user2017.pdf)

The `bssm` package was originally developed with the support of Academy of 
Finland grants 284513, 312605, and 311877. Current development is focused on usability and improvidevelopemnt is 

## Installation

You can install the released version of bssm from [CRAN](https://CRAN.R-project.org) with:

```{r}
install.packages("bssm")
```

And the development version from [GitHub](https://github.com/) with:

```{r}
# install.packages("devtools")
devtools::install_github("helske/bssm")
```
Or from R-universe with 
  
```{r}
install.packages("bssm", repos = "https://helske.r-universe.dev")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(bssm)
## basic example code
```

## Recent changes (For all changes, see NEWS file.)
  
#### bssm 1.1.6 (Release date: 2021-09-06)
  
  * Cleaned codes and added more comprehensive tests in line with pkgcheck 
  tests. This resulted in finding and fixing multiple bugs:
  * Fixed a bug in EKF-based particle filter which returned filtered estimates 
  also in place of one-step ahead predictions.
  * Fixed a bug which caused an error in suggest_N for nlg_ssm.
  * Fixed a bug which caused incorrect sampling of smoothing distribution for 
  ar1_lg model when predicting past or when using simulation smoother.
  * Fixed a bug which caused an error when predicting past values in 
  multivariate time series case.
  * Fixed sampling of negative binomial distribution in predict method, which 
  used std::negative_binomial which converts non-integer phi to integer. 
  Sampling now uses Gamma-Poisson mixture for simulation.
  
#### bssm 1.1.4 (Release date: 2021-04-13)
  
  * Better documentation for SV model, and changed ordering of arguments to emphasise the 
  recommended parameterization.
  * Fixed predict method for SV model.
  
#### bssm 1.1.3-2 (Release date: 2021-02-24)
  
  * Fixed missing parenthesis causing compilation fail in case of no OpenMP support.
  * Added pandoc version >= 1.12.3 to system requirements.
  
#### bssm 1.1.3-1 (Release date: 2021-02-22)
  
  * Fixed PM-MCMC and DA-MCMC for SDE models and added an example to `ssm_sde`.
  * Added vignette for SDE models.
  * Updated citation information and streamlined the main vignette.
  
#### bssm 1.1.2 (Release date: 2021-02-08)
  
  * Some bug fixes, see NEWS for details.
  
#### bssm 1.1.0 (Release date: 2021-01-19)
  
  
  * Added function `suggest_N` which can be used to choose 
  suitable number of particles for IS-MCMC.
  * Added function `post_correct` which can be used to update 
  previous approximate MCMC with IS-weights.
  * Gamma priors are now supported in easy-to-use models such as `bsm_lg`. 
  * The adaptation of the proposal distribution now continues also after the burn-in by default. 
  * Changed default MCMC type to typically most efficient and robust IS2.
  * Renamed `nsim` argument to `particles` in most of the R functions (`nsim` also works with a warning).
  * Fixed a bug with bsm models with covariates, where all standard deviation parameters were fixed. 
  This resulted error within MCMC algorithms.
  * Fixed a dimension drop bug in the predict method which caused error for univariate models.
  * Fixed few typos in vignette (thanks Kyle Hussman) and added more examples.

#### bssm 1.0.1-1 (Release date: 2020-11-12)
  
  
  * Added an argument `future` for predict method which allows 
  predictions for current time points by supplying the original model 
  (e.g., for posterior predictive checks). 
  At the same time the argument name `future_model` was changed to `model`.
  * Fixed a bug in summary.mcmc_run which resulted error when 
  trying to obtain summary for states only.
  * Added a check for Kalman filter for a degenerate case where all 
  observational level and state level variances are zero.
  * Renamed argument `n_threads` to `threads` for consistency 
  with `iter` and `burnin` arguments.
  * Improved documentation, added examples.
  * Added a vignette regarding psi-APF for non-linear models.
  
#### bssm 1.0.0 (Release date: 2020-06-09)
  
  Major update
  
  * Major changes for model definitions, now model updating and priors 
  can be defined via R functions (non-linear and SDE models still rely on C++ snippets).
  * Added support for multivariate non-Gaussian models.
  * Added support for gamma distributions.
  * Added the function as.data.frame for mcmc output which converts the MCMC samples 
  to data.frame format for easier post-processing.
  * Added truncated normal prior.
  * Many argument names and model building functions have been changed for clarity and consistency.
  * Major overhaul of C++ internals which can bring minor efficiency gains and smaller installation size.
  * Allow zero as initial value for positive-constrained parameters of bsm models.
  * Small changes to summary method which can now return also only summaries of the states.
  * Fixed a bug in initializing run_mcmc for negative binomial model. 
  * Fixed a bug in phi-APF for non-linear models.
  * Reimplemented predict method which now always produces data frame of samples.
  
#### bssm 0.1.11 (Release date: 2020-02-25)
  
  * Switched (back) to approximate posterior in RAM for PM-SPDK and PM-PSI, 
  as it seems to work better with noisy likelihood estimates.
  * Print and summary methods for MCMC output are now coherent in their output.
  
#### bssm 0.1.10 (Release date: 2020-02-04)
  
  * Fixed missing weight update for IS-SPDK without OPENMP flag.
  * Removed unused usage argument ... from expand_sample.
  
#### bssm 0.1.9 (Release date: 2020-01-27)
  
  * Fixed state sampling for PM-MCMC with SPDK.
  * Added ts attribute for svm model.
  * Corrected asymptotic variance for summary methods.
  
  For older versions, see NEWS file.
  
