title: "Logistic growth model with bssm"
author: |
| Jouni Helske
| University of Jyväskylä, Department of Mathematics and Statistics, Finland
date: "13 February 2017"
link-citations: true
output: 
pdf_document:
fig_caption: yes
fig_crop: no
fig_height: 6
fig_width: 8
vignette: |
%\VignetteIndexEntry{Logistic growth model with bssm}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
%\VignetteDepends{ggplot2}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Logistic growth model

This vignette shows how to model general non-linear state space models with `bssm`. The general non-linear Gaussian model in `bssm` has following form:

$$
y_t = Z(t, \alpha_t, \theta, \psi, \psi_t) + H(t, \alpha_t, \theta, \psi, \psi_t)\epsilon_t,\\
\alpha_{t+1} = T(t, \alpha_t, \theta, \psi, \psi_t) + R(t, \alpha_t, \theta, \psi, \psi_t)\eta_t,\\
\alpha_1 \sim N(a_1(\theta, \psi), P_1(\theta, \psi)),
$$
with $t=1,\ldots, n$, $\epsilon_t ~ N(0,\textrm{I}_p)$, and $\eta ~ N(0,\textrm{I}_k)$. Here vector $\theta$ contains the unknown model parameters, $\psi$ contains known time-invariant parameters, and $\psi_t$ consists of $n \times r$ matrix of time-varying parameters. Functions $T(\cdot)$, $H(\cdot)$, $T(\cdot)$, $R(\cdot)$,$a_1(\cdot)$, $P_1(\cdot)$, as well as functions defining the Jacobians of $Z(\cdot)$ and $T(\cdot)$ and the prior distribution for $\theta$ must be defined by user as a external pointers to `C++` functions, which can sound intimidating at first, but is actually pretty simple, as this vignette hopefully illustrates.

As an example, consider a logistic growth model

Let's first simulate some data:
```{r data}
set.seed(1)

#parameters
K <- 100 # carrying capacity
p0 <- 10 # population size at t = 0
r <- .2 # growth rate

#sample time
dT <- .1

#observation times
t <- seq(0.1, 25, dT)

# simulate true population size (=p) at the observation times
p <- K * p0 * exp(r * t) / (K + p0 * (exp(r * t) - 1))

# observations
y <- p + rnorm(length(t), 0, 5)
```

## Model in bssm

The functions determining the model functions are given in file `model_functions.cpp`. Here are few pieces from the file. The first one defines the 
state transition function $T(\cdot)$:

```{Rcpp, eval = FALSE}
arma::vec T_fn(const unsigned int t, const arma::vec& alpha, 
  const arma::vec& theta, const arma::vec& known_params, 
  const arma::mat& known_tv_params) {
  
  double dT = known_params(0);
  double k = known_params(1);
  
  arma::vec alpha_new(2);
  alpha_new(0) = alpha(0);
  alpha_new(1) = k * alpha(1) * exp(alpha(0) * dT) / 
    (k + alpha(1) * (exp(alpha(0) * dT) - 1));
  
  return alpha_new;
}
```

Let's make less than optimal initial guess for $\theta$, the standard deviation of observational level noise, the standard deviations of the process noises (which are zero but let's pretend that we are not certain about the underlying model), and define the prior distribution for $\alpha_1$:
```{r theta}
initial_theta <- c(3, 0.5, 0.5)

# dT, K, a1 and the prior variances
psi <- c(dT, 100, 0.3, 5, 4, 10)
```

The name of this function does not matter, but it should always return Armadillo vector (`arma::vec`), and have same signature (ie. the order and types of the function's parameters) should always be like above, even though some of the parameters were not used in the body of the function. For details of using Armadillo, see [Armadillo documentation](http://arma.sourceforge.net/docs.html). After defining the appropriate model functions, the `cpp` file should also contain a function for creating external pointers for the aforementioned functions. Why this is needed is more technical issue, but fortunately you can just copy the function from the example file without any modifications. 

After creating the file for `C++` functions, you need to compile the file using `Rcpp`:

```{r pointers}
Rcpp::sourceCpp("growth_model_functions.cpp")
pntrs <- create_xptrs()
```

This takes a few seconds. If you have used line `// [[Rcpp::export]]` before the model functions, you can now test that the functions work as intended:

```{r test}
T_fn(0, c(100, 200), initial_theta, psi, matrix(1))
```

Now the actual model:
```{r model}
library("bssm")
model <- nlg_ssm(y = y, a1=pntrs$a1, P1 = pntrs$P1, 
  Z = pntrs$Z_fn, H = pntrs$H_fn, T = pntrs$T_fn, R = pntrs$R_fn, 
  Z_gn = pntrs$Z_gn, T_gn = pntrs$T_gn,
  theta = initial_theta, log_prior_pdf = pntrs$log_prior_pdf,
  known_params = psi, known_tv_params = matrix(1),
  n_states = 2, n_etas = 2)
```

Let's first run Extended Kalman filter and smoother using our initial guess for $\theta$:
```{r ekf}
out_filter <- ekf(model)
out_smoother <- ekf_smoother(model)
ts.plot(cbind(y, out_filter$att[, 2], out_smoother$alphahat[, 2]), col = 1:3)
ts.plot(cbind(out_filter$att[, 1], out_smoother$alphahat[, 1]), col = 1:2)
```

## Markov chain Monte Carlo

For parameter inference, we can perform full Bayesian inference with \texttt{bssm}. There are multiple choices for the MCMC algorithm in the package, and here we will use $\psi$-PF based pseudo-marginal MCMC with delayed acceptance. Let us compare this approach with EKF-based approximate MCMC (note that the number of MCMC iterations is rather small):

```{r mcmc}
out_mcmc_pm <- run_mcmc(model, n_iter = 5000, nsim_states = 10, method = "pm", 
  simulation_method = "psi")
out_mcmc_ekf <- run_mcmc(model, n_iter = 5000, method = "ekf")
library("diagis")
alpha_ekf <- weighted_mean(out_mcmc_ekf$alpha, out_mcmc_ekf$counts)
alpha_pm <- weighted_mean(out_mcmc_pm$alpha, out_mcmc_pm$counts)
ts.plot(cbind(y, alpha_pm[, 2], alpha_ekf[, 2]), col = 1:3)
ts.plot(cbind(alpha_pm[, 1], alpha_ekf[, 1]), col = 1:2)
```

Let's make some predictions based on our MCMC run:
```{r pred}
future_model <- model
future_model$y <- ts(rep(NA, 100), start = end(model$y))
out_pred <- predict(out_mcmc_pm, future_model, type = "response")
library("ggplot2")
autoplot(out_pred, y = model$y, plot_median = FALSE)

```

## Appendix

This is the full `growth_model_functions.cpp` file:

```{Rcpp, eval = FALSE}
#include <RcppArmadillo.h>
// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::interfaces(r, cpp)]]

// Function for the prior mean of alpha_1
// [[Rcpp::export]]
arma::vec a1_fn(const arma::vec& theta, const arma::vec& known_params) {
 
  arma::vec a1(2);
  a1(0) = known_params(2);
  a1(1) = known_params(3);
  return a1;
}
// Function for the prior covariance matrix of alpha_1
// [[Rcpp::export]]
arma::mat P1_fn(const arma::vec& theta, const arma::vec& known_params) {
  
  arma::mat P1(2, 2, arma::fill::zeros);
  P1(0,0) = known_params(4);
  P1(1,1) = known_params(5);
  return P1;
}

// Function for the observational level standard deviation
// [[Rcpp::export]]
arma::mat H_fn(const unsigned int t, const arma::vec& alpha, const arma::vec& theta, 
  const arma::vec& known_params, const arma::mat& known_tv_params) {
  arma::mat H(1,1);
  H(0, 0) = theta(0);
  return H;
}

// Function for the Cholesky of state level covariance
// [[Rcpp::export]]
arma::mat R_fn(const unsigned int t, const arma::vec& alpha, const arma::vec& theta, 
  const arma::vec& known_params, const arma::mat& known_tv_params) {
  arma::mat R(2, 2, arma::fill::zeros);
  R(0, 0) = theta(1);
  R(1, 1) = theta(2);
  return R;
}


// Z function
// [[Rcpp::export]]
arma::vec Z_fn(const unsigned int t, const arma::vec& alpha, const arma::vec& theta, 
  const arma::vec& known_params, const arma::mat& known_tv_params) {
  arma::vec tmp(1);
  tmp(0) = alpha(1);
  return tmp;
}
// Jacobian of Z function
// [[Rcpp::export]]
arma::mat Z_gn(const unsigned int t, const arma::vec& alpha, const arma::vec& theta, 
  const arma::vec& known_params, const arma::mat& known_tv_params) {
  arma::mat Z_gn(1, 2);
  Z_gn(0, 0) = 0.0;
  Z_gn(0, 1) = 1.0;
  return Z_gn;
}

// T function
// [[Rcpp::export]]
arma::vec T_fn(const unsigned int t, const arma::vec& alpha, const arma::vec& theta, 
  const arma::vec& known_params, const arma::mat& known_tv_params) {
  
  double dT = known_params(0);
  double k = known_params(1);

  arma::vec alpha_new(2);
  alpha_new(0) = alpha(0);
  alpha_new(1) = k * alpha(1) * exp(alpha(0) * dT) / 
    (k + alpha(1) * (exp(alpha(0) * dT) -1));
  
  return alpha_new;
}

// Jacobian of T function
// [[Rcpp::export]]
arma::mat T_gn(const unsigned int t, const arma::vec& alpha, const arma::vec& theta, 
  const arma::vec& known_params, const arma::mat& known_tv_params) {
  
  double dT = known_params(0);
  double k = known_params(1);
  
  double tmp = exp(alpha(0) * dT) / 
    std::pow(k + alpha(1) * (exp(alpha(0) * dT) - 1), 2);
  
  arma::mat Tg(2, 2);
  Tg(0, 0) = 1.0;
  Tg(0, 1) = 0;
  Tg(1, 0) = k * alpha(1) * dT * (k - alpha(1)) * tmp;
  Tg(1, 1) = k * k * tmp;
  
  return Tg;
}



// # log-prior pdf for theta
// [[Rcpp::export]]
double log_prior_pdf(const arma::vec& theta) {
  
  double log_pdf;
  if(arma::any(theta < 0)) {
     log_pdf = -arma::datum::inf;
   } else {
    // weakly informative priors. 
    // Note that negative values are handled above
    log_pdf = 2.0 * (R::dnorm(theta(0), 0, 10, 1) + R::dnorm(theta(1), 0, 10, 1) + 
      R::dnorm(theta(2), 0, 10, 1));
  }
  return log_pdf;
}

// Create pointers, no need to touch this if
// you don't alter the function names above
// [[Rcpp::export]]
Rcpp::List create_xptrs() {
  // typedef for a pointer of nonlinear function of model equation returning vec
  typedef arma::vec (*vec_fnPtr)(const unsigned int t, const arma::vec& alpha, const arma::vec& theta, 
    const arma::vec& known_params, const arma::mat& known_tv_params);
  // typedef for a pointer of nonlinear function of model equation returning mat
  typedef arma::mat (*mat_fnPtr)(const unsigned int t, const arma::vec& alpha, const arma::vec& theta, 
    const arma::vec& known_params, const arma::mat& known_tv_params);
  // typedef for a pointer of nonlinear function of model equation returning vec
  typedef arma::vec (*vec_initfnPtr)(const arma::vec& theta, const arma::vec& known_params);
  // typedef for a pointer of nonlinear function of model equation returning mat
  typedef arma::mat (*mat_initfnPtr)(const arma::vec& theta, const arma::vec& known_params);
  // typedef for a pointer of log-prior function
  typedef double (*double_fnPtr)(const arma::vec&);
  
  return Rcpp::List::create(
    Rcpp::Named("a1_fn") = Rcpp::XPtr<vec_initfnPtr>(new vec_initfnPtr(&a1_fn)),
    Rcpp::Named("P1_fn") = Rcpp::XPtr<mat_initfnPtr>(new mat_initfnPtr(&P1_fn)),
    Rcpp::Named("Z_fn") = Rcpp::XPtr<vec_fnPtr>(new vec_fnPtr(&Z_fn)),
    Rcpp::Named("H_fn") = Rcpp::XPtr<mat_fnPtr>(new mat_fnPtr(&H_fn)),
    Rcpp::Named("T_fn") = Rcpp::XPtr<vec_fnPtr>(new vec_fnPtr(&T_fn)),
    Rcpp::Named("R_fn") = Rcpp::XPtr<mat_fnPtr>(new mat_fnPtr(&R_fn)),
    Rcpp::Named("Z_gn") = Rcpp::XPtr<mat_fnPtr>(new mat_fnPtr(&Z_gn)),
    Rcpp::Named("T_gn") = Rcpp::XPtr<mat_fnPtr>(new mat_fnPtr(&T_gn)),
    Rcpp::Named("log_prior_pdf") = 
      Rcpp::XPtr<double_fnPtr>(new double_fnPtr(&log_prior_pdf)));
}

```

